Ornaments as indicators of social change before and after European arrival at Kiwulan, Northeastern Taiwan

### 2019-09-19 analysis/paper/paper.Rmd
shorten the abstract, change the some words to british spelling, and change format of figures to JEPG

---
title: "Ornaments as indicators of social change before and after European arrival at Kiwulan, Northeastern Taiwan"
author:
  - Li-Ying Wang
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  In East Asia, direct European colonial rule on local communities was much less conspicuous than in other parts of the world. In this region, the question of long-lasting indirect impacts on indigenous peoples in the periphery of colonial control remains largely unanswered. Using trade ornaments recovered from Kiwulan (1400-1900 AD) in northeastern Taiwan, I explore the indirect impacts of European colonial activities on local societies. I examine the cultural contexts, variety, and spatial distributions of trade ornaments in three periods relating to different cultural interactions. The results indicate the diversity of ornaments was greater during the European presence period compared to previous periods and their spatial distribution was clustered. This hints at increasing social inequality in a colonial context. This study demonstrates the use of foreign ornaments as a proxy to detect indirect colonial influence on local indigenous populations and inform on social changes relating to trade with colonial groups. Ornaments can give insights into the amplification of social inequality stimulated by European colonization. It also shows the agency of indigenous people to incorporate ornaments into their social system.
keywords: |
  Historical archaeology; Iron Age; European colonization; Trade ornaments; Exchange networks; Taiwan; Monte Carlo simulation
highlights: |
  1. Evidence of the quantity, variety, and spatial distribution of trade ornaments in both local cultural and colonial context is provided.
  2. Results of hypothesis testing point to greater amount and diversity of ornament types with non-random, and most highly concentrated spatial clusters during the European presence.
  3. Trade ornament as a prestige good may result in an increase in social inequality due to competition of wealth accumulation between individuals.
  4. Trade ornament can be a good proxy to detect indirect colonial influence on local indigenous populations and inform on social changes relating to trade with colonial groups.
---

@@

```{r, setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/"
)
# prepare for JAS:R
# https://www.elsevier.com/journals/journal-of-archaeological-science-reports/2352-409x/guide-for-authors#2000
# not exceed 6000 words excluding figures, tables, highlights 
image_width_1_col_inch = 90/25.4 # image Single column	w = 90 mm 
image_width_2_col_inch = 190/25.4 # image Double column w = 190 mm 
```

@@

Taiwan became part of global trade systems during the early 17th century due to western European expansions that connected Europe and Asia. Despite the colonization of Taiwan by the Spanish and Dutch from 1624 to 1662, there has been little discussion of the archaeology of indigenous responses to the European colonization. @Berrocal2018 examined both archaeological and historical records of Heping Dao in northern Taiwan. This was the location of the colony of San Salvador founded by the Spanish in 1626 and taken over by the Dutch in 1642. Berrocal et al. uncovered parts of the foundation of a Spanish church and a European cemetery that indicate the evidence of European occupation. These remains indicate a direct impact on the local indigenous populations through their modification of the local landscape. In contrast to Heping Dao, Yilan in Northeastern Taiwan is a physically isolated area adjacent to colonial centers and there are no traces of European settlement but occasional interactions with the Europeans. Yilan, also called "Kavalan" by local indigenous people, is an alluvial plain circumscribed by the Pacific and mountains on the sides. The rugged Hsuehshan Range to the west and Central Range to the south of the Yilan Plain that made it not easily accessible to the European colonists and the colonial control was weaker compared to other parts of this region [@Andrade2007; @Kang2012]. This makes Yilan an especially unique location for exploring indirect colonial effects on local indigenous settlements living beyond the borders of direct colonial occupation.

@@

Chinese historical records from 1829, 1837, and 1852 during the Qing dynasty [1616-1911] contain some notes on the purposes of ornaments from Yilan [@Chen1963, pp. 228, 308; @Ke1993, pp. 11, 126; @Yao1996, pp. 77]. According to those records, indigenous people in Yilan wore ornaments in ceremonial contexts to display their wealth and status. Among those ornaments, golden fish-shaped necklaces made of copper threads had high value due to their delicacy and the materials invested in production. These were usually possessed by wealthy people. Other people wore agate beads or glass beads on their head or neck to participate in ceremonies. In 1895, at the beginning of Japanese colonization, a field survey for plains indigenous groups reported that golden necklaces were not used in Yilan at that time, but elderly people still used beads [@Ino1996, pp. 227-232].

@@

# Excavations at Kiwulan in northeastern Taiwan
```{r, KWL-map, fig.cap = "Map showing the location of Kiwulan, and other places in northern Taiwan named in the text. Map data from naturalearthdata.com", fig.width=image_width_2_col_inch}
# need to add the locations of Keelung, Tamsui, and hoping dao
library(here)
knitr::include_graphics(here("analysis", "figures", "kiwulan-location-map.png"))
```

@@

ggsave(plot = KWL_excavation_map, 
       here("analysis", "figures", "KWL-excavation-map.png"), width = 190, units = "mm")

knitr::include_graphics(here("analysis", "figures", "KWL-excavation-map.png"))
```

Kiwulan (Figure \@ref(fig:KWL-map)) is located at northern Yilan and was a rescue archaeology project that carried out from 2001 to 2004 in advance of water diversion project and construction of a road bridge. The total area includes eight open area sections of 262 excavation squares (4 m by 4 m) reaching 3,814 m^2^ in total [@Chen2007]. The archaeological evidence includes a large amount of artifacts, burials, middens, post holes, wooden pillars, and stone structures, all of which indicates it was a settlement. Most ornaments were found in situ while digging and some were collected through screens with 2 mm and 1.5 mm mesh. The ornaments studied in this paper were found from 40 adjacent squares in the largest open area, located in the middle part of the excavation, see Figure \@ref(fig:KWL-sam-area). They were sampled because those units were stratigraphically intact with the least disturbance by modern construction activity, compared to excavation squares on the periphery of the site.

@@

European colonization ended in 1662 when they were defeated by the kingdom of Tungning founded by Koxinga from China. Later in 1683, the Qing dynasty ruled over Taiwan and a large wave of Han Chinese migrated to Yilan during the late 18th century. The archaeological signature of the Chinese period at Kiwulan is the large amount and varieties of Chinese porcelains, clay stoves, and a few opium pipe-bowls that were identified as items commonly used in the daily life of Chinese and the use of bricks and tiles for architectural buildings [@Hsieh2009]. Also, the description of the Chinese migration to Yilan can be seen from official Chinese records written in the early 19th century [@Chen1963; @Ke1993].

@@

```{r plot-all-ornaments-count, fig.cap= "Frequency of the major class of ornaments at Kiwulan. Frequency represents artifact counts", fig.width=image_width_1_col_inch}
# need to reorder, try https://forcats.tidyverse.org/reference/index.html
# reorder factor for categories (by frequency) and period (manually)
# total number of ornaments for each categories 
ornaments_count <-
  ornaments_period %>% 
  filter(Categories != "Unknown Metal") %>%
  ungroup() %>% # need to upgroup because of rowwise or factor in previous chunk?
  mutate(Categories = fct_lump(Categories, 
                               n = 5)) %>% 
  group_by(Categories) %>%
  count() %>% 
  ungroup() %>% 
  mutate(Categories = fct_reorder(Categories, n))
plot_ornaments_count <- 
ggplot(ornaments_count,
       aes(Categories, n)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(limits = c(0, 160)) +
  labs(y = "Frequency",
       x = "Categories") +
  theme_minimal(base_size = 10)
ggsave(plot = plot_ornaments_count,
       here("analysis", "figures",
       "plot-ornaments-count.png"),
       width = 90, 
       height = 40,
       units = "mm")
knitr::include_graphics(here("analysis", 
                             "figures", 
                             "plot-ornaments-count.png"))
```

```{r tidy-data-for-table}
# clean some 
ornaments_period_clean <- 
ornaments_period %>% 
  mutate(Type = str_remove(Type, "prism/"),
         Type = ifelse(Type == "glass bead", "medium bead", Type),
         Type = ifelse(Type == "small I", "plain small", Type),
         Type = ifelse(Type == "small II", "thin small", Type),
         Type = ifelse(Type == "large thin string", "thin large", Type))
ornaments_shape <-
  ornaments_period_clean %>% 
  group_by(period, Categories, Type) %>% 
  tally() %>% 
  spread(period, n, fill = 0) %>% 
  filter(!Categories %in% c("Unknown Metal", 
                            "Perforated plate", 
                            "Knited item",
                            "Hair pin",
                            "Bone Ornament",
                            "Metal Piece",
                            "Glass ring")) %>%
  filter(!Type %in% c("other",
                      "unclassified")) %>% 
  arrange(Categories, desc(`European Presence`))
# making table
knitr::kable(ornaments_shape,
             caption = "Ornament subtype at Kiwulan. The numbers represent artifact counts ")
# some basic stat
sam_post_hole <- nrow(ornaments_period)
total_post_hole <- nrow(kwl_upper)
sample_per <- round(sam_post_hole/total_post_hole*100, 1)
```

@@

Ornaments from 40 sampling squares were excavated in-situ in different archaeological contexts, including 406 pieces from post-hole areas, 3173 pieces from 30 burials, and 27 pieces from middens. Burials unearthed a large amount of ornaments in the form of bead strings that consisted of hundreds to thousands of pieces. The ornaments from the burial context are not analyzed here since not many burials can be assigned to a time period with confidence. In this paper I focus on the post hole areas where `r sam_post_hole` ornaments were recovered, which accounts for `r sample_per` % of the total number from the same context in the Upper Layer Culture at Kiwulan. Post hole areas represents domestic spaces suitable for testing for changes in the patterns of distribution relating to social organization. Figure \@ref(fig:plot-all-ornaments-count) shows that the most common ornament is golden bead with the number of 143, followed by 121 glass beads, 46 agate beads, 35 metal rings, and 22 metal bells (see also Figure 4). The variety of subtype and their individual frequency are summarized in Table \@ref(tab:tidy-data-for-table) in terms of three time episodes.

@@

To enable re-use of materials and improve reprodicibility and transparency [@Marwick2017], the entire R code [@Rlanguage2019] used for all the analysis and visualizations contained in this paper is included in DOI 10.17605/OSF.IO/R8YGA. Also in this version-controlled compendium [@Marwick2018] are the raw data for all the tests reported here. All of the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this repository. The code is released under the MIT license, the data as CC-0, and figures as CC-BY, to enable maximum re-use.

```{r, plot-counts-by-period, fig.cap = "Frequency of the major ornament across different time periods."}
### delete one plot
# categories of ornaments by period
ornaments_count <-
  ornaments_period %>% 
  filter(!is.na(`6-layers`),
         Categories != "Unknown Metal") %>%
  ungroup() %>% 
  mutate(Categories = fct_lump(Categories, 
                               n = 5)) %>% 
  mutate(Categories = fct_infreq(Categories)) %>% 
  mutate(Categories = fct_rev(Categories)) %>% 
  group_by(period, Categories) %>%
  count()
# histogram facetting for three periods
plot_ornaments_count_three_periods <- 
ggplot(ornaments_count) +
  geom_col(aes(Categories, n)) +
  facet_wrap(~ period) +
  coord_flip() +
  labs(x = "Categories",
       y = "Frequency") +
  theme_minimal(base_size = 12)

ggsave(plot = plot_ornaments_count_three_periods,
       here("analysis", "figures",
       "plot-ornaments-count-three-periods.png"),
       width = 190, 
       height = 80,
       units = "mm")

knitr::include_graphics(here("analysis", 
                             "figures", 
                             "plot-ornaments-count-three-periods.png"))
```

@@

```{r agate-bead-fre-plot}
# plot for different shape of agate bead by period
agate_bead_shape <-
  ornaments_shape_count %>% 
  filter(Categories == 'Agate bead',
         Type != 'other')
base_size_multi_plot <- 10
agate_bead_shape_plot <- 
ggplot(agate_bead_shape) +
  geom_col(aes(reorder(Type, n), n)) +
  facet_wrap(~ period) +
  coord_flip() +
  labs(y = "",
       x = "Agate bead") +
  theme_minimal(base_size = base_size_multi_plot) +
  theme(strip.background = element_blank(), 
        strip.text = element_blank())
```

@@

```{r plot-altogether, fig.cap= "Frequency of ornament subtypes showing the changes in frequency across time periods for metal rings, agate beads, bells, and glass beads."}
library(cowplot)
plot_multi_types_by_period <- 
plot_grid(ring_shape_plot,
          agate_bead_shape_plot, 
          bell_shape_plot, 
          glass_bead_plot,
          ncol=1,
          align = "h",
          axis = "lr",
          rel_widths = 1,
          rel_heights = c(1,1,0.7,0.62))
ggsave(plot = plot_multi_types_by_period,
       here("analysis", "figures",
       "plot-multi-types-by-period.png"),
       width = 190, 
       height = 150,
       units = "mm")
knitr::include_graphics(here("analysis", 
                             "figures", 
                             "plot-multi-types-by-period.png"))
```

```{r regression}
# tidy up the data for correlation testing
ornaments_diversity_test <-
  ornaments_shape_count %>%
  group_by(period) %>% 
  mutate(sam_size = sum(n)) %>% 
  add_count() %>% 
  select(period, n, sam_size) %>% 
  distinct()
ornaments_diversity_test_2 <-
  ornaments_shape_count %>%
  unite(diversity, c(period, Categories), sep = "-", remove = FALSE) %>% 
  group_by(diversity) %>% 
  mutate(sam_size = sum(n)) %>% 
  add_count() %>% 
  select(diversity, n, sam_size) %>% 
  separate(diversity, c("period","subtype"), sep = "-", remove = FALSE) %>% 
  distinct()
# spearman test
spearman_result <- 
cor.test( ~ n + sam_size, 
         data = ornaments_diversity_test_2,
         method = "spearman",
         continuity = FALSE,
         conf.level = 0.95)
spearman_result_stat <- round(spearman_result$statistic, 2)
spearman_result_rho <- unname(round(spearman_result$estimate, 2))
spearman_result_p_value <- round(spearman_result$p.value, 2)
```

@@

# plot spatial pattern for all ornaments
plot_spatial_distribution_all_ornaments_by_period <- 
ggplot() +
  geom_sf(data = sample_units, fill = NA) +
  geom_sf(data = ornaments_period_join, 
          aes(geometry = geometry, 
              fill =n), 
          alpha = 0.9) +
  facet_wrap(~ period) +
  labs(fill = "frequency") +
  scale_fill_viridis(direction = -1) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        strip.text.x = element_text(size = 11))

ggsave(plot = plot_spatial_distribution_all_ornaments_by_period,
       here("analysis", "figures",
       "plot-spatial-distribution-all-ornaments-by-period.png"),
       width = 190, 
       height = 50,
       units = "mm")
knitr::include_graphics(here("analysis", 
                             "figures", "plot-spatial-distribution-all-ornaments-by-period.png"))
```

@@

plot_spatial_distribution_some_ornaments_by_period <-
ggplot() +
  geom_sf(data = sample_units, fill = NA) +
  geom_sf(data = ornaments_period_join_selected_types, 
          aes(geometry = geometry,
              fill = n), alpha = 0.9) +
  facet_grid(rows = vars(Categories),
             cols = vars(period)) +
  labs(title = "",
       fill = "frequency") +
  scale_fill_viridis(direction = -1) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        strip.text.x = element_text(size = 8))

ggsave(plot = plot_spatial_distribution_some_ornaments_by_period,
       here("analysis", "figures",
       "plot-spatial-distribution-some-ornaments-by-period.png"),
       width = 190, 
       height = 130,
       units = "mm")

knitr::include_graphics(here("analysis", 
                             "figures", "plot-spatial-distribution-some-ornaments-by-period.png"))
```

@@

```{r transformation-sptial-data}
# transform data to PPP format and sp SpatialPoints
library(spatstat)
library(maptools)
library(raster)
#store x and y coords in two vectors, and create two vectors xrange and yrange with dimensions of triangle that contain all points, by period
# use common range for each period
lon_all <- ornaments_period_join_cen_all$lon
lat_all <- ornaments_period_join_cen_all$lat
xrange <- range(lon_all, na.rm=T)
yrange <- range(lat_all, na.rm=T)
# create a function to apply to each period
density_analysis <- function(x, ...) {
  # create ppp and sp(spacial points) format
lon_tmp <- x$lon
lat_tmp <- x$lat
ornaments_point_ppp_tmp <- ppp(lon_tmp, lat_tmp, xrange, yrange)
ornaments_point_sp_tmp <- as(ornaments_point_ppp_tmp, "SpatialPoints")
# kernel density estimation with contour
kernel_tmp <- density(ornaments_point_ppp_tmp)
kernel_to_plot <- 
  plot(kernel_tmp, main = "")
  contour(kernel_tmp, add = TRUE)
  title(..., line = -2, adj = 0.4)
return(list(ornaments_point_ppp = ornaments_point_ppp_tmp,
            kernel_to_plot = kernel_to_plot))
}
```

>>>>

---
title: "Ornaments as indicators of social change before and after European arrival at Kiwulan, Northeastern Taiwan"
author:
  - Li-Ying Wang
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/antiquity.csl" # Insert path for the bib-style
abstract: |
  Long-lasting indirect impacts on indigenous peoples in the periphery of colonial control are poorly understood, especially in East Asia. Trade ornaments from Kiwulan (1400-1900 AD) in northeastern Taiwan show the indirect impacts of European colonial activities on local societies. The diversity of ornaments was greater during the period of European presence compared to previous periods, and their spatial distribution was more clustered. This hints at increasing social inequality resulting from a colonial influence. Ornaments give insights into the increasing social inequality stimulated by European colonisation, and show the agency of indigenous people to incorporate ornaments into their social system.
keywords: |
  Historical archaeology; Iron Age; European colonisation; Trade ornaments; Exchange networks; Taiwan; Monte Carlo simulation
highlights: |
  1. Evidence of the quantity, variety, and spatial distribution of trade ornaments in both local cultural and colonial context is provided.
  2. Results of hypothesis testing point to greater amount and diversity of ornament types with non-random, and most highly concentrated spatial clusters during the European presence.
  3. Trade ornament as a prestige good may result in an increase in social inequality due to competition of wealth accumulation between individuals.
  4. Trade ornament can be a good proxy to detect indirect colonial influence on local indigenous populations and inform on social changes relating to trade with colonial groups.
---

@@

```{r, setup, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  comment = "#>",
  fig.path = "../figures/"
)
# prepare for Antiquity https://antiquity.ac.uk/submit-paper
# should be no more than 13.5cm wide and at a minimum resolution of 300dpi. 
# not exceed 6000 words excluding figures, tables, highlights 
image_width_1_col_inch = 90/25.4 # image Single column	w = 90 mm 
image_width_2_col_inch = 190/25.4 # image Double column w = 190 mm 
```

@@

Taiwan became part of global trade systems during the early 17th century due to western European expansions that connected Europe and Asia. Despite the colonisation of Taiwan by the Spanish and Dutch from 1624 to 1662, there has been little discussion of the archaeology of indigenous responses to the European colonisation. @Berrocal2018 examined both archaeological and historical records of Heping Dao in northern Taiwan. This was the location of the colony of San Salvador founded by the Spanish in 1626 and taken over by the Dutch in 1642. Berrocal et al. uncovered parts of the foundation of a Spanish church and a European cemetery that indicate the evidence of European occupation. These remains indicate a direct impact on the local indigenous populations through their modification of the local landscape. In contrast to Heping Dao, Yilan in Northeastern Taiwan is a physically isolated area adjacent to colonial centres and there are no traces of European settlement but occasional interactions with the Europeans. Yilan, also called "Kavalan" by local indigenous people, is an alluvial plain circumscribed by the Pacific and mountains on the sides. The rugged Hsuehshan Range to the west and Central Range to the south of the Yilan Plain that made it not easily accessible to the European colonists and the colonial control was weaker compared to other parts of this region [@Andrade2007; @Kang2012]. This makes Yilan an especially unique location for exploring indirect colonial effects on local indigenous settlements living beyond the borders of direct colonial occupation.

@@

Chinese historical records from 1829, 1837, and 1852 during the Qing dynasty [1616-1911] contain some notes on the purposes of ornaments from Yilan [@Chen1963, pp. 228, 308; @Ke1993, pp. 11, 126; @Yao1996, pp. 77]. According to those records, indigenous people in Yilan wore ornaments in ceremonial contexts to display their wealth and status. Among those ornaments, golden fish-shaped necklaces made of copper threads had high value due to their delicacy and the materials invested in production. These were usually possessed by wealthy people. Other people wore agate beads or glass beads on their head or neck to participate in ceremonies. In 1895, at the beginning of Japanese colonisation, a field survey for plains indigenous groups reported that golden necklaces were not used in Yilan at that time, but elderly people still used beads [@Ino1996, pp. 227-232].

@@

# Excavations at Kiwulan in northeastern Taiwan
```{r, KWL-map, fig.cap = "Map showing the location of Kiwulan, and other places in northern Taiwan named in the text. Map data from naturalearthdata.com", fig.width=image_width_2_col_inch}
# need to add the locations of Keelung, Tamsui, and hoping dao
library(here)
knitr::include_graphics(here("analysis", "figures", "kiwulan-location-map.jpg"))
```

@@

ggsave(plot = KWL_excavation_map, 
       here("analysis", "figures", "KWL-excavation-map.jpg"), 
       width = 135,
       dpi = 300,
       units = "mm")

knitr::include_graphics(here("analysis", "figures", "KWL-excavation-map.jpg"))
```

Kiwulan (Figure \@ref(fig:KWL-map)) is located at northern Yilan and was a rescue archaeology project that carried out from 2001 to 2004 in advance of water diversion project and construction of a road bridge. The total area includes eight open area sections of 262 excavation squares (4 m by 4 m) reaching 3,814 m^2^ in total [@Chen2007]. The archaeological evidence includes a large amount of artefacts, burials, middens, post holes, wooden pillars, and stone structures, all of which indicates it was a settlement. Most ornaments were found in situ while digging and some were collected through screens with 2 mm and 1.5 mm mesh. The ornaments studied in this paper were found from 40 adjacent squares in the largest open area, located in the middle part of the excavation, see Figure \@ref(fig:KWL-sam-area). They were sampled because those units were stratigraphically intact with the least disturbance by modern construction activity, compared to excavation squares on the periphery of the site.

@@

European colonisation ended in 1662 when they were defeated by the kingdom of Tungning founded by Koxinga from China. Later in 1683, the Qing dynasty ruled over Taiwan and a large wave of Han Chinese migrated to Yilan during the late 18th century. The archaeological signature of the Chinese period at Kiwulan is the large amount and varieties of Chinese porcelains, clay stoves, and a few opium pipe-bowls that were identified as items commonly used in the daily life of Chinese and the use of bricks and tiles for architectural buildings [@Hsieh2009]. Also, the description of the Chinese migration to Yilan can be seen from official Chinese records written in the early 19th century [@Chen1963; @Ke1993].

@@

```{r plot-all-ornaments-count, fig.cap= "Frequency of the major class of ornaments at Kiwulan. Frequency represents artefact counts", fig.width=image_width_1_col_inch}
# need to reorder, try https://forcats.tidyverse.org/reference/index.html
# reorder factor for categories (by frequency) and period (manually)
# total number of ornaments for each categories 
ornaments_count <-
  ornaments_period %>% 
  filter(Categories != "Unknown Metal") %>%
  ungroup() %>% # need to upgroup because of rowwise or factor in previous chunk?
  mutate(Categories = fct_lump(Categories, 
                               n = 5)) %>% 
  group_by(Categories) %>%
  count() %>% 
  ungroup() %>% 
  mutate(Categories = fct_reorder(Categories, n))
plot_ornaments_count <- 
ggplot(ornaments_count,
       aes(Categories, n)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(limits = c(0, 160)) +
  labs(y = "Frequency",
       x = "Categories") +
  theme_minimal(base_size = 10)
ggsave(plot = plot_ornaments_count,
       here("analysis", "figures",
       "plot-ornaments-count.jpg"),
       width = 90, 
       height = 40,
       dpi = 300,
       units = "mm")
knitr::include_graphics(here("analysis", 
                             "figures", 
                             "plot-ornaments-count.jpg"))
```

```{r tidy-data-for-table}
# clean some 
ornaments_period_clean <- 
ornaments_period %>% 
  mutate(Type = str_remove(Type, "prism/"),
         Type = ifelse(Type == "glass bead", "medium bead", Type),
         Type = ifelse(Type == "small I", "plain small", Type),
         Type = ifelse(Type == "small II", "thin small", Type),
         Type = ifelse(Type == "large thin string", "thin large", Type))
ornaments_shape <-
  ornaments_period_clean %>% 
  group_by(period, Categories, Type) %>% 
  tally() %>% 
  spread(period, n, fill = 0) %>% 
  filter(!Categories %in% c("Unknown Metal", 
                            "Perforated plate", 
                            "Knited item",
                            "Hair pin",
                            "Bone Ornament",
                            "Metal Piece",
                            "Glass ring")) %>%
  filter(!Type %in% c("other",
                      "unclassified")) %>% 
  arrange(Categories, desc(`European Presence`))
# making table
knitr::kable(ornaments_shape,
             caption = "Ornament subtype at Kiwulan. The numbers represent artefact counts ")
# some basic stat
sam_post_hole <- nrow(ornaments_period)
total_post_hole <- nrow(kwl_upper)
sample_per <- round(sam_post_hole/total_post_hole*100, 1)
```

@@

Ornaments from 40 sampling squares were excavated in-situ in different archaeological contexts, including 406 pieces from post-hole areas, 3173 pieces from 30 burials, and 27 pieces from middens. Burials unearthed a large amount of ornaments in the form of bead strings that consisted of hundreds to thousands of pieces. The ornaments from the burial context are not analyzed here since not many burials can be assigned to a time period with confidence. In this paper I focus on the post hole areas where `r sam_post_hole` ornaments were recovered, which accounts for `r sample_per` % of the total number from the same context in the Upper Layer Culture at Kiwulan. Post hole areas represents domestic spaces suitable for testing for changes in the patterns of distribution relating to social organisation. Figure \@ref(fig:plot-all-ornaments-count) shows that the most common ornament is golden bead with the number of 143, followed by 121 glass beads, 46 agate beads, 35 metal rings, and 22 metal bells (see also Figure 4). The variety of subtype and their individual frequency are summarized in Table \@ref(tab:tidy-data-for-table) in terms of three time episodes.

@@

To enable re-use of materials and improve reproducibility and transparency [@Marwick2017], the entire R code [@Rlanguage2019] used for all the analysis and visualisations contained in this paper is included in http://doi.org/10.17605/OSF.IO/R8YGA. Also in this version-controlled compendium [@Marwick2018] are the raw data for all the tests reported here. All of the figures, tables, and statistical test results presented here can be independently reproduced with the code and data in this repository. The code is released under the MIT license, the data as CC-0, and figures as CC-BY, to enable maximum re-use.

```{r, plot-counts-by-period, fig.cap = "Frequency of the major ornament across different time periods."}
### delete one plot
# categories of ornaments by period
ornaments_count <-
  ornaments_period %>% 
  filter(!is.na(`6-layers`),
         Categories != "Unknown Metal") %>%
  ungroup() %>% 
  mutate(Categories = fct_lump(Categories, 
                               n = 5)) %>% 
  mutate(Categories = fct_infreq(Categories)) %>% 
  mutate(Categories = fct_rev(Categories)) %>% 
  group_by(period, Categories) %>%
  count()
# histogram facetting for three periods
plot_ornaments_count_three_periods <- 
ggplot(ornaments_count) +
  geom_col(aes(Categories, n)) +
  facet_wrap(~ period) +
  coord_flip() +
  labs(x = "Categories",
       y = "Frequency") +
  theme_minimal(base_size = 10)

ggsave(plot = plot_ornaments_count_three_periods,
       here("analysis", "figures",
       "plot-ornaments-count-three-periods.jpg"),
       width = 135, 
       dpi = 300,
       height = 50,
       units = "mm")

knitr::include_graphics(here("analysis", 
                             "figures", 
                             "plot-ornaments-count-three-periods.jpg"))
```

@@

```{r agate-bead-fre-plot}
# plot for different shape of agate bead by period
agate_bead_shape <-
  ornaments_shape_count %>% 
  filter(Categories == 'Agate bead',
         Type != 'other')
base_size_multi_plot <- 8
agate_bead_shape_plot <- 
ggplot(agate_bead_shape) +
  geom_col(aes(reorder(Type, n), n)) +
  facet_wrap(~ period) +
  coord_flip() +
  labs(y = "",
       x = "Agate bead") +
  theme_minimal(base_size = base_size_multi_plot) +
  theme(strip.background = element_blank(), 
        strip.text = element_blank())
```

@@

```{r plot-altogether, fig.cap= "Frequency of ornament subtypes showing the changes in frequency across time periods for metal rings, agate beads, bells, and glass beads."}
library(cowplot)
plot_multi_types_by_period <- 
plot_grid(ring_shape_plot,
          agate_bead_shape_plot, 
          bell_shape_plot, 
          glass_bead_plot,
          ncol=1,
          align = "h",
          axis = "lr",
          rel_widths = 1,
          rel_heights = c(1,1,0.7,0.62))
ggsave(plot = plot_multi_types_by_period,
       here("analysis", "figures",
       "plot-multi-types-by-period.jpg"),
       width = 135, 
       dpi = 300,
       height = 95,
       units = "mm")
knitr::include_graphics(here("analysis", 
                             "figures", 
                             "plot-multi-types-by-period.jpg"))
```

```{r regression}
# tidy up the data for correlation testing
ornaments_diversity_test <-
  ornaments_shape_count %>%
  group_by(period) %>% 
  mutate(sam_size = sum(n)) %>% 
  add_count() %>% 
  dplyr::select(period, n, sam_size) %>% 
  distinct()
ornaments_diversity_test_2 <-
  ornaments_shape_count %>%
  unite(diversity, c(period, Categories), sep = "-", remove = FALSE) %>% 
  group_by(diversity) %>% 
  mutate(sam_size = sum(n)) %>% 
  add_count() %>% 
  dplyr::select(diversity, n, sam_size) %>% 
  separate(diversity, c("period","subtype"), sep = "-", remove = FALSE) %>% 
  distinct()
# spearman test
spearman_result <- 
cor.test( ~ n + sam_size, 
         data = ornaments_diversity_test_2,
         method = "spearman",
         continuity = FALSE,
         conf.level = 0.95)
spearman_result_stat <- round(spearman_result$statistic, 2)
spearman_result_rho <- unname(round(spearman_result$estimate, 2))
spearman_result_p_value <- round(spearman_result$p.value, 5)
```

@@

# plot spatial pattern for all ornaments
plot_spatial_distribution_all_ornaments_by_period <- 
ggplot() +
  geom_sf(data = sample_units, fill = NA) +
  geom_sf(data = ornaments_period_join, 
          aes(geometry = geometry, 
              fill =n), 
          alpha = 0.9) +
  facet_wrap(~ period) +
  labs(fill = "frequency") +
  scale_fill_viridis(direction = -1) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        strip.text.x = element_text(size = 6),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 6),
        legend.key.size =  unit(3, "mm"))

ggsave(plot = plot_spatial_distribution_all_ornaments_by_period,
       here("analysis", "figures",
       "plot-spatial-distribution-all-ornaments-by-period.jpg"),
       width = 135, 
       height = 31,
       dpi = 300,
       units = "mm")
knitr::include_graphics(here("analysis", 
                             "figures", "plot-spatial-distribution-all-ornaments-by-period.jpg"))
```

@@

plot_spatial_distribution_some_ornaments_by_period <-
ggplot() +
  geom_sf(data = sample_units, fill = NA) +
  geom_sf(data = ornaments_period_join_selected_types, 
          aes(geometry = geometry,
              fill = n), alpha = 0.9) +
  facet_grid(rows = vars(Categories),
             cols = vars(period)) +
  labs(title = "",
       fill = "frequency") +
  scale_fill_viridis(direction = -1) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        strip.text.x = element_text(size = 6))

ggsave(plot = plot_spatial_distribution_some_ornaments_by_period,
       here("analysis", "figures",
       "plot-spatial-distribution-some-ornaments-by-period.jpg"),
       width = 135, 
       height = 82,
       dpi = 300,
       units = "mm")

knitr::include_graphics(here("analysis", 
                             "figures", "plot-spatial-distribution-some-ornaments-by-period.jpg"))
```

@@

```{r transformation-sptial-data}
# transform data to PPP format and sp SpatialPoints
library(spatstat)
library(maptools)
library(raster)
#store x and y coords in two vectors, and create two vectors xrange and yrange with dimensions of triangle that contain all points, by period
# use common range for each period
lon_all <- ornaments_period_join_cen_all$lon
lat_all <- ornaments_period_join_cen_all$lat
xrange <- range(lon_all, na.rm=T)
yrange <- range(lat_all, na.rm=T)
# create a function to apply to each period
density_analysis <- function(x, ...) {
  # create ppp and sp(spacial points) format
lon_tmp <- x$lon
lat_tmp <- x$lat
ornaments_point_ppp_tmp <- ppp(lon_tmp, lat_tmp, xrange, yrange)
ornaments_point_sp_tmp <- as(ornaments_point_ppp_tmp, "SpatialPoints")
# kernel density estimation with contour
kernel_tmp <- density(ornaments_point_ppp_tmp)
kernel_to_plot <- 
  plot(kernel_tmp, main = "")
  contour(kernel_tmp, add = TRUE)
  title(..., line = -0.3, adj = 0.4, cex.main = 0.8)
return(list(ornaments_point_ppp = ornaments_point_ppp_tmp,
            kernel_to_plot = kernel_to_plot))
}
```

```{r plot-all-densities, fig.cap= "Kernel density map for ornaments by periods. Used the bandwidth based on Silverman (1986)'s rule of thumb."}
# apply function to each data frame in the list
titles <- names(ornaments_period_join_cen)
jpeg(
  here("analysis", "figures",
       "plot-kde-maps.jpg"), 
    width = 135,
    height = 40,
    units = "mm",
  res = 300)
par(mfrow = c(1, 3),
    oma = c(1,1,1,1) + 0.1,
    mar = c(0,0,1,1) + 0.1)
kernels_for_plotting <- 
  map2(ornaments_period_join_cen, 
       titles,
      ~density_analysis(.x, main = .y))
invisible(dev.off())
knitr::include_graphics(here("analysis", 
                             "figures", 
                             "plot-kde-maps.jpg"))
```
